Gammapy: A Python package for gamma-ray astronomy

### 2023-01-20 src/ms-review.tex
Remov code stats table as it duplicates info from pie chart

\section{Gamma-ray Data Analysis}
\label{sec:gammaray-data-analysis}
%
\begin{figure*}[ht!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/irfs.pdf}
	\caption{
		Example instrument response functions of some experiments and
        observatories for which \gammapy can analyse data. The \cta IRFs
        are from the prod5 production. The \hess IRFs are from the DL3 DR1,
        using observation ID 033787. The point spread function shows the 65\&
        containment radius of the PSF. The \fermi IRFs are from \textit{pass8}.
        \todo{Add more instruments? HAWC? MAGIC?}
    }
	\label{fig:irfs}
\end{figure*}
%
% This might not be the bast place compared to introduction
The data reduction process in \gammaray astronomy is usually split into two parts.
The first one deals with the data processing from camera measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This sequence, sometimes referred to as low-level analysis, is usually very specific to
a given observation technique and even to a given instrument.

@@

\subsection{Data analysis workflow}

In Gammapy, the handling of DL3 data is performed by classes and methods
in the \code{gammapy.data} (see Section~\ref{ssec:gammapy-data}) and the \code{gammapy.irf}
(see Section~\ref{ssec:gammapy-irf}) subpackages.

The first step in the analysis is the selection and extraction of observations
based of their metadata including information such as pointing direction, observation
time and observation conditions.

The next step of the analysis is the data reduction, where all observation events and instrument
responses are filled into or projected onto a common physical coordinate system, defined by
a map geometry. The definition of the map geometry typically consists of a spectral dimension
defined by a binned energy axis and of spatial dimensions, which either define 
a spherical projection from celestial coordinates to a pixelised image space
(for 3-dimensional or cube analysis) or a region on the sky (for regular spectral analysis).
The \code{gammapy.maps} subpackage provides general multidimensional geometry objects
(\code{Geom}) and the associated data structures (\code{Maps}) (see Section~\ref{ssec:gammapy-maps}).

@@

The counts data and the reduced IRFs in the form of maps are bundled into datasets
that represent the data level 4 (DL4). Those reduced datasets can be written to disk,
in a format specific to \gammapy to allow users to read them back at any time later
for modeling and fitting. Different variations of such datasets support different 
analysis methods and fit statistics (e.g. Poisson statistics with known background or
with OFF background measurements). The datasets are used to perform joint-likelihood
fitting allowing to combine different measurements, e.g. from different observations
but also from different instruments or event classes. They can also be used for binned
simulation as well as event sampling to simulate DL3 events data.
The various DL4 objects and the associated functionalities are
implemented in the \code{gammapy.datasets} subpackage (see Section~\ref{ssec:gammapy-datasets}).

@@

\section{\gammapy Package}
\label{sec:gammapy-package}
\subsection{Overview}
\label{ssec:overview}
%
\begin{figure*}[ht!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the groups for the different data levels and reduction steps
        from raw gamma-like events on the left, to high level science products
        on the right. The direction of the data flow is illustrated with the
        grey arrows. The gray folder icons represent the different sub-packages
        in \gammapy and their names. Below each icon there is a list of the most
        important objects defined in the sub-package.
    }
	\label{fig:data_flow}
\end{figure*}
%
The \gammapy package is structured into multiple sub-packages. The definition
of the content of the different sub-packages follows mostly the stages in the
data reduction workflow described in the previous section. Sub-packages
either contain data structures representing data at different data reduction
levels or contain algorithms to transition between the different data reduction
levels.

@@

\subsection{gammapy.irf}
\label{ssec:gammapy-irf}
%
%\begin{figure}
%	%	\import{code-examples/generated/}{gp_makers}
%
%	\caption{In Figure~\ref{ig*:minted:irf_examples}, we show some example of Aeff,
%		PSF, Edisp and BKG read and plotted from a typical \irf.}
%	\label{ig*:minted:irf_examples}
%\end{figure}
%
The \code{gammapy.irf} sub-package contains all classes and functionality
to handle IRFs in a variety of formats.
Usually, \irfs store instrument properties in the form of multi-dimensional
tables, with quantities expressed in terms of energy (true or reconstructed),
off-axis angles or cartesian detector coordinates. The main information stored in
the common \gammaray \irfs are the effective area, energy dispersion,
point spread function and background rate. The \code{gammapy.irf}
sub-package can open and access specific \irf extensions,
interpolate and evaluate the quantities of interest on both energy and spatial
axes, convert their format or units in different kinds, plot or write them into
output files. In the following, we list the main classes of the
sub-package:

@@

\subsubsection{Background model}
The backgropund rate can be represented in \gammapy as either a 2-dimensional
data structure (\code{Background2D}) of count rate normalised per steradians and energy at different
reconstructed energies and offset angles or as rate per steradians and energy, as a
function of reconstructed energy and detector coordinates (\code{Background3D}).
In the former, the background is expected to follow a radially symmetric shape,
while in the latter, it can be more complex.

@@

\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/cta_galactic_center.pdf}
	\caption{
		Example of 1D spectral analysis of the Galactic Center for three simulated CTA
		observations for the 1DC dataset. The left image shows the maps of counts with the measurement
		region in white and background regions overlaid in different colors. The right image
		shows the resulting spectral points and their corresponding log-likelihood
		profiles.} \label{fig:cta_galactic_center} \end{figure*}

\subsection{3D Analysis}
\label{ssec:3d-analysis}
The 1D analysis approach is a powerful tool to measure the spectrum of an
isolated source. However, more complicated situations require a more careful
treatment. In a field of view containing several overlapping sources, the 1D
approach cannot disentangle the contribution of each source to the total flux in
the selected region. Sources with extended or complex morphology can result in
the measured flux being underestimated, and heavily dependent on the choice of
extraction region. Additionally, the 1D approach neglects the energy-dependence
of the PSF.

@@

\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/cube_analysis.pdf}
	\caption{Example of a 3D analysis for simulated sources with point-like, Gaussian
		and shell-like morphology. The simulation uses \textit{prod5} \irfs from \cta.
		The left image shows a significance map where the three simulated sources can be
		seen. The middle figure shows another significance map, but this time after
		subtracting the best-fit model for each of the sources, which are displayed in
		black. The right figure shows the contribution of each source model to the
		circular region of radius 0.5\textdegree drawn in the left image, together with
		the excess counts inside that region. } \label{fig:cube_analysis}
\end{figure*}

@@

\begin{table}
	\import{tables/generated/}{codestats}
	\caption{
        Overview of used programming languages and distribution of code across the different file
        categories in the \gammapy code base. The most right column list the total number of lines
        in a file. The \textit{comment} column lists the number of comments in the files
        (including method and class docstrings), the \textit{blank} column lists the
        number of blank lines in the file. The uppermost rows distinguish between
        code that implements actual functionality and code the implements tests.
        All documentataion in \gammapy is implemented as dosctrings, \text{reStructuredText} (RST)
        files or notebooks.
    }
	\label{table:codestats:data}
\end{table}

\begin{figure}[t]
	\centering
	\includegraphics[width=0.5\textwidth]{figures/codestats.pdf}
	\caption{
		Overview of used programming languages and distribution of code across the different file
        categories in the \gammapy code base.
    } \label{fig:codestats:lang}
\end{figure}

>>>>

\section{Gamma-ray Data Analysis}
\label{sec:gammaray-data-analysis}
% This might not be the bast place compared to introduction
The data reduction process in \gammaray astronomy is usually split into two parts.
The first one deals with the data processing from camera measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This sequence, sometimes referred to as low-level analysis, is usually very specific to
a given observation technique and even to a given instrument.

@@

\subsection{Data analysis workflow}
%
\begin{figure*}[h!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the groups for the different data levels and reduction steps
        from raw gamma-like events on the left, to high level science products
        on the right. The direction of the data flow is illustrated with the
        grey arrows. The gray folder icons represent the different sub-packages
        in \gammapy and their names. Below each icon there is a list of the most
        important objects defined in the sub-package.
    }
	\label{fig:data_flow}
\end{figure*}
%

The first step in \gammaray data analysis is the selection and extraction of observations
based of their metadata including information such as pointing direction, observation
time and observation conditions. The access to the events data and instrument 
reponse per observation is supported by classes and methods
in the \code{gammapy.data} (see Section~\ref{ssec:gammapy-data}) and the \code{gammapy.irf}
(see Section~\ref{ssec:gammapy-irf}) subpackages.

The next step of the analysis is the data reduction, where all observation events and instrument
responses are filled into or projected onto a common physical coordinate system, defined by
a map geometry. The definition of the map geometry typically consists of a spectral dimension
defined by a binned energy axis and of spatial dimensions, which either define 
a spherical projection from celestial coordinates to a pixelised image space
or a single region on the sky. The \code{gammapy.maps} subpackage provides
general multidimensional geometry objects and the associated data structures
(see Section~\ref{ssec:gammapy-maps}).

@@

The counts data and the reduced IRFs in the form of maps are bundled into datasets
that represent the fourth data level (DL4). Those reduced datasets can be written to disk,
in a format specific to \gammapy to allow users to read them back at any time later
for modeling and fitting. Different variations of such datasets support different 
analysis methods and fit statistics. The datasets are used to perform joint-likelihood
fitting allowing to combine different measurements, e.g. from different observations
but also from different instruments or event classes. They can also be used for binned
simulation as well as event sampling to simulate DL3 events data.
The various DL4 objects and the associated functionalities are
implemented in the \code{gammapy.datasets} subpackage (see Section~\ref{ssec:gammapy-datasets}).

@@

\section{\gammapy Package}
\label{sec:gammapy-package}
\subsection{Overview}
\label{ssec:overview}
%
%
The \gammapy package is structured into multiple sub-packages. The definition
of the content of the different sub-packages follows mostly the stages in the
data reduction workflow described in the previous section. Sub-packages
either contain data structures representing data at different data reduction
levels or contain algorithms to transition between the different data reduction
levels.

@@

\subsection{gammapy.irf}
\label{ssec:gammapy-irf}
%
%
\begin{figure*}[ht!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/irfs.pdf}
	\caption{
		Using \code{gammapy.irf} to read and plot instrument repsonse functions.
		Here we show example instrument response functions of some experiments and
        observatories for which \gammapy can analyse data. The \cta IRFs
        are from the prod5 production. The \hess IRFs are from the DL3 DR1,
        using observation ID 033787. The point spread function shows the 65\&
        containment radius of the PSF. The \fermi IRFs are from \textit{pass8}.
    }
	\label{fig:irfs}
\end{figure*}
%
%
The \code{gammapy.irf} sub-package contains all classes and functionality
to handle IRFs in a variety of formats.
Usually, \irfs store instrument properties in the form of multi-dimensional
tables, with quantities expressed in terms of energy (true or reconstructed),
off-axis angles or cartesian detector coordinates. The main information stored in
the common \gammaray \irfs are the effective area, energy dispersion,
point spread function and background rate. The \code{gammapy.irf}
sub-package can open and access specific \irf extensions,
interpolate and evaluate the quantities of interest on both energy and spatial
axes, convert their format or units in different kinds, plot or write them into
output files. In the following, we list the main classes of the
sub-package:

@@

\subsubsection{Instrumental Background}
The instrumental backgropund rate can be represented in \gammapy as either a 2-dimensional
data structure (\code{Background2D}) of count rate normalised per steradians and energy at different
reconstructed energies and offset angles or as rate per steradians and energy, as a
function of reconstructed energy and detector coordinates (\code{Background3D}).
In the former, the background is expected to follow a radially symmetric shape,
while in the latter, it can be more complex.

Some example IRFs that are read and plotted with \gammapy are show in Figure~\ref{fig:irfs}.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.\textwidth]{figures/cta_galactic_center.pdf}
	\caption{
		Example of 1D spectral analysis of the Galactic Center for three simulated CTA
		observations for the 1DC dataset. The left image shows the maps of counts with the measurement
		region in white and background regions overlaid in different colors. The right image
		shows the resulting spectral points and their corresponding log-likelihood
		profiles.} \label{fig:cta_galactic_center} \end{figure*}

\subsection{3D Analysis}
\label{ssec:3d-analysis}
%
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/cube_analysis.pdf}
	\caption{Example of a 3D analysis for simulated sources with point-like, Gaussian
		and shell-like morphology. The simulation uses \textit{prod5} \irfs from \cta.
		The left image shows a significance map where the three simulated sources can be
		seen. The middle figure shows another significance map, but this time after
		subtracting the best-fit model for each of the sources, which are displayed in
		black. The right figure shows the contribution of each source model to the
		circular region of radius 0.5\textdegree drawn in the left image, together with
		the excess counts inside that region. } \label{fig:cube_analysis}
\end{figure*}
%
The 1D analysis approach is a powerful tool to measure the spectrum of an
isolated source. However, more complicated situations require a more careful
treatment. In a field of view containing several overlapping sources, the 1D
approach cannot disentangle the contribution of each source to the total flux in
the selected region. Sources with extended or complex morphology can result in
the measured flux being underestimated, and heavily dependent on the choice of
extraction region. Additionally, the 1D approach neglects the energy-dependence
of the PSF.

@@

@@

% \begin{table}
% 	\import{tables/generated/}{codestats}
% 	\caption{
%         Overview of used programming languages and distribution of code across the different file
%         categories in the \gammapy code base. The most right column list the total number of lines
%         in a file. The \textit{comment} column lists the number of comments in the files
%         (including method and class docstrings), the \textit{blank} column lists the
%         number of blank lines in the file. The uppermost rows distinguish between
%         code that implements actual functionality and code the implements tests.
%         All documentataion in \gammapy is implemented as dosctrings, \text{reStructuredText} (RST)
%         files or notebooks.
%     }
% 	\label{table:codestats:data}
% \end{table}

\begin{figure}[t]
	\centering
	\includegraphics[width=0.5\textwidth]{figures/codestats.pdf}
	\caption{
		Overview of used programming languages and distribution of code across the different file
        categories in the \gammapy code base. The total number of lines is $\approx 50 000$.
    } \label{fig:codestats:lang}
\end{figure}
